---
title: "The ensurer package"
author: "Stefan Holst Milton Bache"
date: "November, 2014"
output: html_document
---

<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{The ensurer}
-->

The ensurer package
===================

*Stefan Holst Milton Bache. September 2014.*

# Introduction
The `R` programming language is becoming an integrated part of data solutions
and analyses in many production environments. This is no surprise: it is very 
powerful and goals are often very quickly accomplished with `R` compared to many 
of its competitors, and it is therefore a primary tool of choice for many 
statisticians, data scientists, and the like.

A common argument against it, however, is the lack of e.g. type safety, and 
a compiler that guards against many potential code problems. 
To make the situation even 
more dangerous, it is very common in `R` to have functions that accept different
types for the same argument, and they may even return different types depending
on various inputs and circumstances. 
The very large community, which is one of `R`'s strengths,
should also raise caution, as developers adhere to different principles, styles,
and standards.

When putting an `R` program into production one should therefore be careful
and aware of the various sources of risk. 
External data sources or schematics may change; 
packages on which you depend may change; resources may be temporarily 
unavailable, etc. While there is little hope to solve these kinds of problems
altogether, it can be essential *become aware* of issues as soon as they arise
and take proper action. In particular, unexpected behavior may not 
itself raise an error (due to the lack of type safety) 
but may result in "corrupted" data which may propagate,
and when errors occur down the road, the initial source of the problem can be
hard and/or time consuming to track down.

The `ensurer` package has one aim: to make it as simple as possible to
ensure expected/needed/desired properties of your data, and to take proper 
action upon failure to comply.

# Basics
The `ensurer` package does not depend on other packages, but its semantics are
designed with `%>%`, the `magrittr` pipe, in mind. This will make it very 
natural to attach an ensuring contract to the result of a data pipeline 
before assigning it to a name, or returning it from a function, etc.
The examples below will use `%>%` but it should be clear how to proceed without.

Usage is simple. There are only two functions to remember:

```{r, eval=FALSE}
1. ensure_that:  function(value., ..., fail_with, err_desc)  [short-hand alias: ensure]
2. ensures_that: function(..., fail_with, err_desc)          [short-hand alias: ensures]
```

The first (in imperative form) takes a value (`value.`) and a set of 
conditions (`...`) which are verified for the value. Upon success, 
the value itself is returned and upon
failure an error is raised (default behavior, which can be modified.)
The second (in present tense) takes only the conditions and creates a 
function which can be reused to ensure these. The arguments `fail_with` and 
`err_desc` can be (but need not be) specified to alter the default behavior.

Stating the conditions is simple, each is an expression which result in 
`TRUE` or `FALSE` and they are separated with commas (technical note: conditions
are checked with `isTRUE`, and conditions should therefore not result in vectors
as this will count as failure even if all entries are `TRUE`.) 
To reference the value itself, use the dot-placeholder (`.`). 

As an example, suppose you have a function `get_matrix` which is supposed to
return a square matrix. The function may read an external text file, or 
otherwise be vulnerable to data corruption. Suppose that you need the 
matrix to be square, and you need it to be numeric. Here's how:

```{r, eval=FALSE}
the_matrix <-
  get_matrix() %>%
  ensure_that(is.numeric(.), 
              NCOL(.) == NROW(.))
```

If `get_matrix` returns valid data, everything is fine. But suppose
a character somehow found its way in, coercing everything to the character type.
In this case the default behavior is to raise an error with details on which
conditions failed (all will be tested, even if previous ones fail). 
In this case the error is:

```{r eval=FALSE}
 Error: conditions failed for call 'get_matrix %>% ensu .. NCOL(.) == NROW(.))':
   * is.numeric(.)
``` 

The second function, `ensures_that` (note: present tense, not imperative form), 
is ideal for creating reusable contracts,
so that the same conditions need not be specified several places with similar
purpose. Using the above example, if several matrices need to be square and
numeric, do:

```{r eval=FALSE}
ensure_square_numeric <- 
  ensures_that(NCOL(.) == NROW(.),
               is.numeric(.))

m1 <- get_matrix()       %>% ensure_square_numeric
m2 <- get_other_matrix() %>% ensure_square_numeric
```

Note how the present tense form, "ensure**s** that", and the imperative form, 
"ensure that" makes the statements very much like regular sentences.

It is straight forward to combine contracts made with `ensures_that` with 
on-the-fly additional conditions:

```{r eval=FALSE}
m3 <-
  get_matrix() %>%
  ensure_square_numeric %>%
  ensure_that(all(. < 10))
```

However, note that conditions stated in the same `ensure*` call are all 
checked, i.e. in the following example the error received is not the same:

```{r eval=FALSE}
# Only the first error is recorded.
letters %>%
  ensure_that(length(.) == 10) %>%
  ensure_that(all(. == toupper(.)))

# Both errors are recorded.
letters %>%
  ensure_that(length(.) == 10,
              all(. == toupper(.)))
```

To round off this section, here is how the first example is made without 
the pipe operator, `%>%`:

```{r eval=FALSE}
the_matrix <- 
  ensure_that(get_matrix(), 
      NCOL(.) == NROW(.),
      is.numeric(.))
```

# Naming arguments
It is possible to assign names for use in the conditions, such that no
external variable declaration is needed for additional information needed
in the conditions. This can be used to make the code more readable, or
to avoid the same computation twice. To assign a value in the `ensure(s)_that`
call, simply use named arguments (and avoid `fail_with` and `err_desc`):
```{r, eval = FALSE}
some_object <- 
  some_computation() %>%
  ensure_that(foo(a) == bar(a), a = baz(x, y, z))
```

# Adding existing contracts to new contracts
Sometimes it is useful to have a set of contracts for different purposes,
and combine them in situations where more than one of them apply.
As described above, one can chain together multiple `ensure_that` 
statements, but in that situation `ensurer` will not check all conditions
from all contracts. Another option is to add existing contracts, constructed
with `ensures_that` to a new `ensure(s)_that` call. To let `ensurer` 
know that the argument is an existing contract to be added, use a unary 
`+` operator:
```{r, eval = FALSE}
matrix_is_square <- ensures_that(NROW(.) == NCOL(.))
all_positive     <- ensures_that(all(. > 0))


matrix(runif(16), 4, 4) %>% 
  ensure(+matrix_is_square, +all_positive)
```
Any named arguments in an added contract will also be available in the
new contract.

# Customizing individual condition messages
To make the description of each of the failed conditions more user-friendly
and readable, `ensurer` lets you specify a description to conditions
where the code is not transparent enough. To do this use a formula as
condition: `condition ~ "message if fails"`. As an example:
```{r, eval = FALSE}
ensure_character <- 
  ensures_that(is.character(.) ~ "vector must be of character type.")

1:10 %>% ensure_character

Error: conditions failed for call '1:10 %>% ensure_character':
   * vector must be of character type. 
```

# Ensuring function return values
The `ensures_that` function also provides a useful mechanism for ensuring 
the characteristics of function return values. This will both make functions 
safer, but will also provide users with specific knowledge about what they
can expect of your functions. Here are a few pseudo examples:

```{r eval=FALSE}
`: numeric` <- ensures_that(is.numeric(.))

get_quote <- function(ticker) `: numeric`({
  # some code that extracts latest quote for the ticker
})


`: square matrix` <- ensures_that(is.numeric(.),
                                  is.matrix(.),
                                  NCOL(.) == NROW(.))
                                  
var_covar <- function(vec) `: square matrix`({
  # some code that produces a variance-covariance matrix.
})                                 
```

While the naming convention used above is not necessary, it is expressive
when used for this purpose; but there is nothing different about the 
contracts produced by `ensures_that`. Of course, another more standard option 
is to ensure the return value at the end of the function, yet this won't 
be as transparent to the user of the function.

# Controlling failure
The main purpose of the `ensurer` is to ensure that values are as they are
expected, and if not alert the stakeholder(s) immediately. The default
behavior is therefore to raise an error. There could be several reasons 
to overrule this default, e.g. you may wish to send an email on error, or
you might want to accept some default value if the desired value is not
available.

The `ensurer` has a simple mechanism for changing the behavior by 
specifying the argument `fail_with` to `ensure_that` or 
`ensures_that`. It is possible to pass in a static value which will be returned
(e.g. an empty `data.frame` with the correct columns, or maybe even just
`NA`). More often, however, more dynamic behavior is needed, and for 
this on passes a function of a single argument, which will be applied
to a `simpleError` object (the one that is raised by default):

```{r eval=FALSE}
square_failure <- function(e)
{
	# suppose you had an email function:
	email("maintainer@company.com", subject = "error", body = e$message)
	stop(e)
}

m1 <- 
  get_matrix() %>% 
  ensure_that(
      NCOL(.) == NROW(.),
      is.numeric(.),
      fail_with = square_failure)

m2 <-
  get_matrix() %>% 
  ensure_that(
      NCOL(.) == NROW(.),
      is.numeric(.),
      fail_with = diag(10))
```

In some instances you may want to use the value itself in the error handler,
although you don't know anything about it, except that it does not comply
with the conditions. 
Therefore this should be done with care, and currently there is not direct
access to it (as you may need to think twice before doing it.) In the above 
example, the `square_failure` does not know about `.`, although you can 
fetch it using `get`:

```{r eval=FALSE}
square_failure <- function(e)
{
  # fetch the dot.
  . <- get(".", parent.frame())
  
  # compose a message detailing also the class of the object returned.
  msg <- 
    sprintf("Here is what I know:\n%s\nValue class: %s.",
            e$message, 
            class(.) %>% paste(collapse = ", ")) # there could be several.
  
  # suppose you had an email function:
  email("maintainer@company.com", subject = "error", body = msg)
  
  stop(e)
}
```

It is also possible to use the dot, `.`, directly in anonymous error handlers
defined directly in the call to e.g. `ensure_that`.

One can add a description to the error message without having to specify 
an error function, which can be useful if the same conditions are used 
several places in your code, or simply to add some information to the 
user about what could be the cause of the problem. This is done by 
specifying the named argument `err_desc`.
For example, `RODBC`'s `sqlQuery` returns a `data.frame` on success
and a character string on failure. Suppose you have a function which makes use
of this to fetch some results, and you wish to make your function safe using
the `ensurer`:

```{r eval=FALSE}
`: sql result` <- 
    ensures_that(is.data.frame(.),
                 err_desc = "SQL error.")
                 
daily_results <- function(day) `: sql result`({
    sql <- sprintf("SELECT * FROM RESULTS WHERE DAY = '%s'",
                   format(as.Date(day)))
    ch <- RODBC::odbcDriverConnect("connection_string")
    on.exit(RODBC::odbcClose(ch))
    RODBC::sqlQuery(ch, sql)
})
```

The error would look like this:
```{r, eval=FALSE}
daily_results("2014-10-01")

 Error: conditions failed for call 'daily_results("2014-10-01")':
   * is.data.frame(.)
 Description: SQL error. 
```

# Building on the ensurer functions
In this section it is shown how simple extensions to the functionality 
provided by the `ensurer` can be used to make flexible, yet strict,
safety mechanisms.

A common task is to load data from one or more external sources, 
say from a SQL database, process the data, combine it various ways, and then
produce some result or analysis. This may even be bundled up in a package and
used by someone else who is unaware of the underlying data sources.
If a connection is broken, a schema has changed, or if somehow the data does 
not come out as expected, the software will probably suffer from both
lack of results and some useless error messages. One step towards avoid such
a scenario is to ensure that all data that are pulled into the application are
as expected; i.e. provide some "type safety" to the data objects, 
say `data.frame`s.

Often you can gain a lot of security in your code if you can specify a 
template for risky objects which specifies the necessary details of such objects.
An obvious example is a `data.frame`: you may need to require specific names 
and types of the columns and/or that it is non-empty. Maybe you have several
important datasets that each are specific, yet all require the same type 
of safety. Here is an example of how to specify a "template" and an ensuring 
function that compares an object to such a template.

```{r eval=FALSE}
iris_template <- 
  data.frame(
    Sepal.Length = numeric(0),
    Sepal.Width  = numeric(0),
    Petal.Length = numeric(0),
    Petal.Width  = numeric(0),
    Species      = 
        factor(numeric(0), levels = c("setosa", "versicolor", "virginica"))
)

ensure_as_template <- function(x, tpl)
  ensure_that(x,
    is.data.frame(.),
    identical(class(.), class(tpl)),
    identical(sapply(., class), sapply(tpl, class)),
    identical(sapply(., levels), sapply(tpl, levels))
  )

iris %>% ensure_as_template(iris_template)
```
The `ensure_as_template` is general enough, to accept any other `data.frame`
with a corresponding template. Packages that exposes functions which return
`data.frames` could define such templates (internally or externally) and 
ensure data validity this way.


